---
title: "Activity monitoring data"
author: "Saiprasad Kamath"
date: "Thursday, November 13, 2014"
output: html_document
---

# Reproducible Research: Peer Assessment 1

## Introduction
``` {r setoptions, echo=FALSE}

## opts_chunk$set(echo=TRUE,  cache=TRUE, fig.width=12, fig.height=8, warning=FALSE, message=FALSE)

```

Data analysis of [Activity Monitoring Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip>)

**Source**

The data for this assignment can be downloaded from the course web site:

* **Dataset:** [Activity Monitoring Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip>) [52K]
The variables included in this dataset are:

* **steps:** Number of steps taking in a 5-minute interval (missing values are coded as NA)

* **date:** The date on which the measurement was taken in YYYY-MM-DD format

* **interval:** Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

**Read Data**

```{r readData, echo=TRUE}
## Loading and preprocessing the data
actData <- read.csv("activity.csv", colClasses = c("numeric", "character", "numeric"))
actData["date"] <- as.Date(actData$date, "%Y-%m-%d")

```

  
  
  
**Q1 What is mean total number of steps taken per day?**  
*For this part of the assignment, missing values in the dataset are ignored.*

```{r meansteps, echo=TRUE}
## What is mean total number of steps taken per day?
sum_spd <- aggregate(steps ~ date, data = actData, sum, na.rm = TRUE)
mean_spd <- mean(sum_spd$steps)
median_spd <- median(sum_spd$steps)

format_mean <- format(mean_spd, digits=2)
format_median <- format(median_spd, digits=2)

```
  
  
Mean steps taken per-day is `r format_mean`  
Median steps taken per-day is `r format_median`
  
**Q2 Make a histogram of the total number of steps taken each day**
  
```{r stepsHist01, echo=TRUE}
## Make a histogram of the total number of steps taken each day
hist(sum_spd$steps, main = "Frequency of Steps Taken Per Day",
     xlab = "Steps",
     ylab="Days", col = "steelblue", breaks=10)
```

  
**Q3 What is the average daily Activity pattern?**  

* A. Time Series plot
```{r activityPattern, echo=TRUE}

## What is the average daily Activity pattern?
series_time <- tapply(actData$steps, actData$interval %% 100 / 5 + actData$interval %/% 100 * 12 + 1, mean, na.rm = TRUE)

max_intrv <- which.max(series_time)
start_hour <- (max_intrv * 5) %/% 60
start_min <- (max_intrv * 5) %% 60

end_hour <- ((max_intrv + 1) * 5) %/% 60
end_min <- ((max_intrv + 1) * 5) %% 60

start_ap <- "AM"
end_ap <- "AM"

if(start_hour > 12) {
  start_hour <- start_hour - 12
  start_ap <- "PM"
}

if(end_hour > 12) {
  end_hour <- end_hour - 12
  end_ap <- "PM"
}

plot(row.names(series_time), series_time, type = "l", xlab = "5-minute Interval", 
     ylab = "Average Steps", main = "Average Steps During 5-minute Time Intervals", 
     col = "red")

```

  
  * Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?  
    
The maximum number of steps is a 5 min interval is: **`r start_hour`:`r start_min` `r start_ap` and `r end_hour`:`r end_min` `r end_ap`**  
  
**Q4 Imputing missing values**  
  
* Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r missingvals, echo=TRUE}

## Inputing missing values
num_NAs <- sum(is.na(actData))

```
  
The total number of missing values is: `r format(num_NAs)`  
  
* **Strategy**  
+ First convert data from tidy format to a wide format  
+ Then compute and replace the null values
+ Finally re-convert the data from wide format to tidy  
* Generate Histogram  

```{r strategy, echo=TRUE, warning=FALSE, message=FALSE}
require(reshape)
require(reshape2)
a_wide <- dcast(actData, interval ~ date, value.var="steps", fill=0)
r_wide <- dcast(actData, interval ~ date, value.var = "steps", fill = rowMeans(a_wide, na.rm = TRUE))
reshaped <- reshape(r_wide, direction = "long", varying=list(names(r_wide)[2:length(names(r_wide))]),
                    v.names=c("steps"), timevar="date", idvar=c("interval"),
                    times=names(r_wide)[2:length(names(r_wide))], new.row.names=1:dim(actData)[1])

sum_spd2 <- aggregate(steps ~ date, data = reshaped, sum, na.rm = TRUE)
hist(sum_spd2$steps, main = "Steps Per Day", xlab = "Daily Steps",
     ylab="Days", col = "steelblue", breaks=10)

```

* Calculate and report the mean and median total number of steps taken per day  

```{r meansteps2, echo=TRUE}

## Calculate mean and median for new data
mean_spd2 <- mean(sum_spd2$steps)
median_spd2 <- median(sum_spd2$steps)

```

   
Then mean steps per day is: `r format(mean_spd2)` and the median steps per day is `r format(median_spd2)`  
  
* What is the impact of imputing missing data  
The mean and median values increased by ~1000 steps when missing data was replaced.  
  
**Q5 Are there differences in activity patterns between weekdays and weekends?

```{r weekdaysends, echo=TRUE, warning=FALSE, message=FALSE}
## Are there differences in actData patterns between weekdays and weekends?
require(lattice)
reshaped["date"] <- as.Date(reshaped$date, "%Y-%m-%d")
reshaped[(weekdays(reshaped$date) %in% c("Saturday", "Sunday")), "TypeOfDay"] <- "Weekends"
reshaped[!(weekdays(reshaped$date) %in% c("Saturday", "Sunday")), "TypeOfDay"] <- "Weekdays"

steps <- aggregate(steps ~ interval + TypeOfDay, data = reshaped, mean)
names(steps) <- c("Interval", "TypeOfDay", "Steps")
steps$Interval <- steps$Interval %% 100 / 5 + steps$Interval %/% 100 * 12 + 1

splot <- xyplot(Steps ~ Interval | TypeOfDay, steps, type = "l", layout = c(1, 2), 
                xlab = "5-minute Interval", ylab = "Average Steps")
update(splot,
       main="Weekend Days Versus Weekdays")


```



  

